Uncertainty propagation in deep neural networks for sparse coding

### 2018-03-10 paper/BayesianLISTA.tex
minor text fixes

Neural networks address the sparse coding problem with the soft thresholding nonlinearity, that shifts current estimates towards zero \citep{gregor2010learning}. In this paper we propose a method to propagate uncertainty through the soft thresholding nonlinearity. At every layer the current distribution of the target vector is represented as a spike-and-slab distribution \citep{mitchell1988bayesian}. We show that it can be effectively combined with Gaussian weights of dense layers. The distribution of the target vector propagated through a dense layer and soft thresholding nonlinearity can then be closely approximated with a distribution from the same family.

@@

The paper is organised as following: first, we review neural networks for sparse coding in section \ref{sec:nn_sc} and present a novel Bayesian neural network in section \ref{sec:bayesian_lista}. We then describe uncertainty propagation to obtain the spike-and-slab distribution for an output of a network in section \ref{sec:fprop} and how probabilistic backpropagation is organised to update distributions of weights in section \ref{sec:backpropagation}. After that, we show the experimental results of the algorithm in section \ref{sec:experiments} and finish with discussion and conclusions in sections \ref{sec:discussion} and \ref{sec:conclusions}.

@@

Matrices $\mathbf{W}$, $\mathbf{S}$ are the parameters that are initialised as in \textsc{ista} and then updated with the backpropagation algorithm. Vectors $\mathbf{c}_l$, $\mathbf{b}$ are intermediate vectors that describe forward propagation. $\boldsymbol\beta_l$ is the current approximation of the target coefficient vector at the layer $l$.

@@

\subsection{\uppercase{Approximation quality}}
\label{sec:approx_quality}
We have used two approximations in forward propagation of uncertainty. First, in lemma \ref{thm:matrix_vector} a Gaussian matrix is multiplied by a spike-and-slab vector and their product is approximated with the Gaussian distribution. Second, in lemma \ref{thm:soft_thresholding} the result of soft thresholding of a Gaussian vector is approximated with the spike-and-slab distribution. In this section we demonstrate that these approximations are close to the real distributions.

@@

\subsection{\uppercase{Predictive performance on synthetic data}}
First, the predictive performance of the proposed Bayesian \textsc{lista} is analysed on synthetic data. We generate $N_\text{train}=500$ sparse coefficients vectors $\boldsymbol\beta_n$ each of size $D = 100$. Coefficients $\boldsymbol\beta$ are generated from the spike-and-slab distribution with truncated slab: each component of $\beta_{nd}$ is zero with probability $0.8$ or is from standard Gaussian distribution without interval $(-0.1, 0.1)$ with probability $0.2$. Truncation of the zeros allows to remove from support pattern values with insignificant contributions \citep{xin2016maximal}. To simulate sparse observations, we have generated the random Gaussian design matrix $\mathbf{X} \in \mathbb{R}^{K \times D}$.  The observations are generated according to the \ref{eq:regression_problem} with zero-mean Gaussian noise with standard deviation $0.5$. $\lambda$ is assumed fixed as $0.1$.

>>>>

Neural networks address the sparse coding problem with the soft thresholding nonlinearity, that shifts current estimates towards zero \citep{gregor2010learning}. In this paper we propose a method to propagate uncertainty through the soft thresholding nonlinearity. At every layer the current distribution of the target vector is represented as a spike and slab distribution \citep{mitchell1988bayesian}. We show that it can be effectively combined with Gaussian weights of dense layers. The distribution of the target vector propagated through a dense layer and soft thresholding nonlinearity can then be closely approximated with a distribution from the same family.

@@

The paper is organised as following: first, we review neural networks for sparse coding in section \ref{sec:nn_sc} and present a novel Bayesian neural network in section \ref{sec:bayesian_lista}. We then describe uncertainty propagation to obtain the spike and slab distribution for an output of a network in section \ref{sec:fprop} and how probabilistic backpropagation is organised to update distributions of weights in section \ref{sec:backpropagation}. After that, we show the experimental results of the algorithm in section \ref{sec:experiments} and finish with discussion and conclusions in sections \ref{sec:discussion} and \ref{sec:conclusions}.

@@

Matrices $\mathbf{W}$, $\mathbf{S}$ are the parameters that are initialised as in \textsc{ista} and then updated with the backpropagation algorithm. Vectors $\mathbf{c}_l$, $\mathbf{b}$ are intermediate vectors that describe forward propagation.

@@

\subsection{\uppercase{Approximation quality}}
\label{sec:approx_quality}
We have used two approximations in forward propagation of uncertainty. First, in lemma \ref{thm:matrix_vector} a Gaussian matrix is multiplied by a spike and slab vector and their product is approximated with the Gaussian distribution. Second, in lemma \ref{thm:soft_thresholding} the result of soft thresholding of a Gaussian vector is approximated with the spike and slab distribution. In this section we demonstrate that these approximations are close to the real distributions.

@@

\subsection{\uppercase{Predictive performance on synthetic data}}
First, the predictive performance of the proposed Bayesian \textsc{lista} is analysed on synthetic data. We generate $N_\text{train}=500$ sparse coefficients vectors $\boldsymbol\beta_n$ each of size $D = 100$. Coefficients $\boldsymbol\beta$ are generated from the spike and slab distribution with truncated slab: each component of $\beta_{nd}$ is zero with probability $0.8$ or is from standard Gaussian distribution without interval $(-0.1, 0.1)$ with probability $0.2$. Truncation of the zeros allows to remove from support pattern values with insignificant contributions \citep{xin2016maximal}. To simulate sparse observations, we have generated the random Gaussian design matrix $\mathbf{X} \in \mathbb{R}^{K \times D}$.  The observations are generated according to the \ref{eq:regression_problem} with zero-mean Gaussian noise with standard deviation $0.5$. $\lambda$ is assumed fixed as $0.1$.
